---
title: "Preparing your data for FINN"
format:
  html:
    toc: true
    toc-depth: 3
vignette: >
  %\VignetteIndexEntry{Preparing your data for FINN}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette is a standalone reference for getting your own forest-inventory
data into FINN. It documents the tables FINN needs, their columns and units, and
the builder pipeline `makeObsData()` → `resolveSiteIDs()` → `makeInitCohorts()`.
Every step runs on a tiny bundled raw example, so you see real output.

```{r}
#| label: libraries
library(FINN)
library(data.table)
```

## What FINN needs

FINN is calibrated against a few tidy tables:

- **`obs_dt`** — demographic responses per site × year × species:
  `siteID, year, ba, dbh, trees, growth, mort, reg, species, species_name`.
  Year starts at 1; the initial state (year 0) is *not* included. NAs are allowed
  in any response. Units: `ba` m², `dbh` cm, `trees` count, `growth` relative
  growth rate, `mort` rate, `reg` trees ha⁻¹.
- **`env_dt`** — `siteID, year, <env vars…>`; must cover every site × year in
  `obs_dt`. Z-scale the variables and keep a scaling table for plotting.
- **`init_trees`** — the initial state used to build the starting cohorts:
  `siteID, patchID, year, species, dbh, trees, living`.
- **`dist_dt`** (optional) — `siteID, year, intensity` (fraction of patches
  disturbed that year).

You assemble these from two raw ingredients: a **tree list** (one row per tree ×
inventory) and a **site environment table**. The rest of this vignette builds the
FINN tables from those two.

## Start from a raw tree list

```{r}
#| label: load-example
# A tiny bundled raw example: a few Oregon FIA sites, 4 patches each, 3 inventories.
tree_dt <- fread(system.file("extdata", "example_tree_dt.csv", package = "FINN"))
str(tree_dt)
```

A raw tree record carries, at minimum: `siteName, patchName, treeName, year`
(calendar year of the inventory), `species_name`, `dbh` (cm), and a `status` of
`new` / `alive` / `dead`. Our example also carries `status_before` (the tree's
status at the previous inventory) and a `complete` flag marking trees from fully
re-measured plots.

## Acquiring raw data (example: US FIA)

Turning a public inventory into the raw tree list above needs external downloads,
so it is shown but not run. For the US FIA program:

```{r}
#| label: acquire-fia
#| eval: false
library(rFIA)
fia <- readFIA(dir = "data/FIA", common = TRUE, states = "OR")
# build tree_dt: dbh = DIA * 2.54 (inch -> cm); status from STATUSCD / PREV_STATUS_CD;
# mort_cause from AGENTCD. See playground/calibration/FIA/FIA_dataprep_vignette.R
```

## Define status flags

FINN's builders key off three logical flags derived from `status`:

```{r}
#| label: status-flags
# reg    = newly recruited this inventory
# mort   = died since the previous inventory
# living = currently alive
tree_dt[, reg    := status == "new"]
tree_dt[, mort   := status == "dead" & status_before %chin% c("alive", "new")]
tree_dt[, living := status %chin% c("alive", "new")]
tree_dt[, .(n = .N), by = .(status, reg, mort, living)]
```

## Build `obs_dt` with `makeObsData()`

`makeObsData()` aggregates the tree list to per-patch demographic responses and
lumps rare species into a single `"other"` class. Each argument controls a
data-quality filter:

```{r}
#| label: make-obs
obs_list <- makeObsData(
  tree_dt           = tree_dt[complete == TRUE],  # only fully re-measured plots
  plotsize          = 0.06,            # ha per patch (scales recruitment to per-ha)
  aggregate_by_site = FALSE,           # keep patch-level resolution
  fix_period_length = 10,              # drop sites whose remeasure gap != 10 yr
  dbh_growth_thresh = c(-10, 50),      # drop sites with implausible dbh change
  NspeciesQuantile  = 0.98,            # keep species covering 98% of stems; rest -> "other"
  minNyears         = 2                # need >= 2 inventories per patch
)
obs_dt <- obs_list$obs_dt
head(obs_dt)
```

Use `Nspecies` (a hard cap) or `NspeciesQuantile` (smallest set covering a stem
fraction) to control how many species you model — everything else becomes
`"other"`, which FINN still tracks as a competitor.

## Environmental data

Acquisition from rasters (e.g. WorldClim) needs external files, so it is shown
but not run; we then z-scale a small bundled env table for real.

```{r}
#| label: env-extract
#| eval: false
library(terra)
pts  <- vect(site_dt, geom = c("x", "y"), crs = "EPSG:4326")
bio1 <- rast("wc2.1_30s_bio_1.tif")
env_dt[, temp := terra::extract(bio1, project(pts, crs(bio1)))[, 2]]
```

```{r}
#| label: env-scale
env_dt   <- fread(system.file("extdata", "example_env_dt.csv", package = "FINN"))
env_vars <- c("temp", "tempmax", "tempmin", "prec", "precseas", "precwarmq")

# Z-scale each variable and KEEP the mean/sd so plots can be back-transformed.
env_scales_dt <- data.table(
  variable = env_vars,
  mean     = sapply(env_dt[, ..env_vars], mean),
  sd       = sapply(env_dt[, ..env_vars], sd)
)
for (v in env_vars) {
  s <- env_scales_dt[variable == v]
  env_dt[[v]] <- (env_dt[[v]] - s$mean) / s$sd
}
env_scales_dt
head(env_dt)
```

`env_dt` must cover every `siteName × year` present in `obs_dt`.

## Resolve IDs across tables

Raw tables use names (`siteName`, `patchName`, `species_name`); FINN works with
integer indices aligned across all tables. `resolveSiteIDs()` does that alignment
and can build the initial cohorts in one pass. Pass the **processed** tree list
`obs_list$tree_dt` (not the raw one): `makeObsData()` returns it with the same
species lumping as `obs_dt`, so the two stay consistent.

```{r}
#| label: resolve-ids
inputs <- resolveSiteIDs(
  tree_dt = obs_list$tree_dt,   # processed tree list (species lumped like obs_dt)
  obs_dt  = obs_dt,
  env_dt  = env_dt,
  createInitCohorts = TRUE
)
# obs_dt / env_dt now carry integer siteID/patchID/species, aligned across tables
head(inputs$obs_dt)
inputs$species_dt
```

## Initial cohorts

With `createInitCohorts = TRUE`, `resolveSiteIDs()` returns a ready `CohortMat`
describing the starting state — its `dims` are `c(sites, patches, cohorts)`:

```{r}
#| label: init-cohorts
init_cohorts <- inputs$initCohorts
init_cohorts
```

To build cohorts yourself from a year-0 tree slice, call `makeInitCohorts()`
directly — it expects `siteID, patchID, species, dbh, treeName` (+ optional
`trees`):

```{r}
#| label: init-cohorts-manual
#| eval: false
init_trees <- inputs$tree_dt[living == TRUE & year == 1]
makeInitCohorts(init_trees, Nspecies = uniqueN(inputs$obs_dt$species))
```

## Checklist before fitting

- **Period length** is consistent across inventories (`fix_period_length`); FINN
  assumes a fixed remeasure interval.
- **The initial state** (the first inventory) is used for `init_trees`, and the
  responses in `obs_dt` start *after* it.
- **`env_dt` covers every `siteID × year`** in `obs_dt` — missing rows break the fit.
- **Units**: `dbh` in cm, `ba` in m², `reg` in trees ha⁻¹.
- **Species IDs** are contiguous integers `1..N`, aligned across tables — always
  go through `resolveSiteIDs()` rather than hand-coding them.

You now have `obs_dt`, `env_dt`, `env_scales_dt`, and `init_cohorts` — the inputs
the **Fitting FINN to forest inventory data** vignette consumes.

```{r}
#| label: session
sessionInfo()
```
