---
title: "Fitting FINN to the US National Forest Inventory (FIA) data for Oregon"
format: html
---

Usually dynamic forest models are parameterized in a complex process involving the translation of process knowledge into mathematical formulas or alrorithms and the determiniation of parameters via systematic (inverse) calibration, manual calibration, and possibly lots of try and error. Identifying all true parameters of a model is impossible but carefully piecing together different modules and components of the model from these methods has created a range of models that simulates somewhat plausible patterns and succession trajectories.

While some models rely on the integration of extremely detailed processes, others rely on empirical models that were fitted to data separately and then put together. 

Here we show how the forest gap model FINN can be parameterized with data only and no prior knowledge on any species properties

# load packages and read data

```{r}
#| echo: false
#| message: false
#| warning: false
library(FINN)
library(torch)
library(data.table)
library(ggplot2)

# the data read in here is created by the script
# "fit-fia-data/FIA_dataprep_vignette.R"
env_scales_dt <- fread("fit-fia-data/env_scales_dt.csv")
tree_dt <- fread("fit-fia-data/tree_dt.csv")
env_unscaled_dt <- fread("fit-fia-data/env_unscaled_dt.csv")
env_dt <- fread("fit-fia-data/env_dt.csv")
env_scales_dt <- fread("fit-fia-data/env_scales_dt.csv")
obs_dt <- fread("fit-fia-data/obs_dt.csv")
full_obs_dt <- fread("fit-fia-data/full_obs_dt.csv")
full_tree_dt <- fread("fit-fia-data/full_tree_dt.csv")
species_dt <- fread("fit-fia-data/species_dt.csv")
```

# Calibration data

Currently FINN can be calibrated on forest inventory that provides information on basal area, number of trees, average dbh, growth, mortality, and regeneration. The specific definition of responses is presented in Pichler & Käber (2025). In this example the values have been preprocessed from the FIA data where the data was queried with the `rFIA` package. Many thanks to Daniel Perret from the USDA for providing the code for downloading and formatting the FIA data from the database.

## Observation table

The observation table of FINN consists of a table that contains the columns
- siteID: unique identifier ranging from 1 to N of sites.
- year: the year of observation starting from 1. Note that the year of the initialization state is not part of observation table
- species: ID of the species ranging from 1 to N of species

All responses are defined as the average over each patch. Within one patch these are defined as.
- ba: sum of basal area of the species within the area of the patch in $m^2$. For multiple patches the average across patches is taken.
- dbh: the average dbh of the species within the area of the patch in cm. For multiple patches the average across patches is taken.
- trees: the number of trees of the species within the area of the patch. For multiple patches the average across patches is taken.
- growth: the relative growth rate for each tree averaged over all trees of a species within the patch.
- mort: the mortality rate for all trees of a species within a patch.
- reg: number of trees that regeneration in trees per ha.

This is the observation table of `r max(obs_dt$siteID)` sites with `r max(obs_dt$species)` species

```{r}
#| label: obs-table
#| 

summary(obs_dt)

```
Note that the table can include NAs for any response.

```{r}
#| label: plot-observations
#| fig.height: 8
#| fig.width: 10

obs_dt_long <- melt(
  full_obs_dt,
  id.vars = c("siteID", "year", "species_name", "species"),
  variable.name = "variable",
  value.name = "value"
)
hist(full_tree_dt$rel_growth)
ggplot(
  obs_dt_long[species %in% 1:3], 
  aes(y = value, x = as.factor(year))) +
  geom_boxplot() +
  facet_wrap(~variable, scales = "free_y") +
  theme_minimal()
```

## Environmental Input data

Any environmental input variable can be used as input. Currently FINN only supports environmental input that matches the temporal and spatial dimensions of the obs_dt. Which means that for each siteID and year in obs_dt there must be at least one entry for each environmental predictor. In a future version FINN will provide options to include inputs of any format.

For this example we use bioclimatic variables that were extracted from worldclim. The environmental input was z scaled. Therefore we apply the backtransformation before visualizing the input

```{r}
#| label: env-data-summary

summary(env_unscaled_dt)
par(mfrow =c(2,3))
for(i in names(env_unscaled_dt)[-c(1,2)]) hist(env_unscaled_dt[[i]], xlab = i, main = NULL)

```


## Initialization State

The initial state usually is the state of the forest in the very first year in a data set. For FINN this must be specified as a cohort object which represents the individual trees as cohorts. Here it is crucial to remove this first state from the observation table. The initial cohort can be created by using the `makeInitCohorts()` function with a table contains rows for each individual tree and columns for 

```{r init-trees}

init_trees <- full_tree_dt[year == 0,.(siteID, patchID, year, species, species_name, dbh, trees, living)]
init_trees[living != T | is.na(living), trees := 0,] # set number of trees of all non living trees to 0
summary(init_trees)
init_cohorts <- FINN::makeInitCohorts(init_trees, Nspecies = max(obs_dt$species))

init_cohorts
```




# Configuration of FINN (the FINN object before fitting)

## Process-FINN

## Hybrid-FINN


```{r}
#| label: load-fittet-model
#| 

# 
# m1 <- torch::torch_load("../data/fia_v12_process_finn.pt")
# 
# modulus_transform <- function(g, lambda) {
#   # g: numeric vector (may include NA)
#   # lambda: scalar power, usually 0 < lambda < 1
#   
#   if (length(lambda) != 1 || !is.finite(lambda)) {
#     stop("lambda must be a single finite number")
#   }
#   if (!is.numeric(g)) {
#     stop("g must be numeric")
#   }
# 
#   out <- rep(NA_real_, length(g))  # preserve NA
#   
#   pos_idx <- !is.na(g) & g >= 0
#   neg_idx <- !is.na(g) & g < 0
#   
#   out[pos_idx] <- g[pos_idx]^lambda
#   out[neg_idx] <- -((-g[neg_idx])^lambda)
#   
#   out
# }
# 
# par(mfrow = c(2,2))
# lambda = 0.4
# br <- seq(-50,50,1)
# hist(tree_dt[abs(dbh_growth) < 50]$dbh_growth, breaks = br, xlim = c(-50,50), main = "untransformiert")
# hist(modulus_transform(tree_dt[abs(dbh_growth) < 50]$dbh_growth/10,lambda)*10, breaks = br, xlim = c(-50,50), main = "transformiert")
# plot(tree_dt[abs(dbh_growth) < 50]$dbh_growth, modulus_transform(tree_dt[abs(dbh_growth) < 50]$dbh_growth/10,lambda)*10)
# abline(a = 0, b = 1)
# plot(tree_dt[abs(dbh_growth) < 50]$dbh_growth, modulus_transform(tree_dt[abs(dbh_growth) < 50]$dbh_growth/10,lambda)*10, ylim = c(-5,5), xlim = c(-5,5))
# abline(a = 0, b = 1)
# 

```


# Understanding the Output (fitted FINN object)

## Parameters
Effect sizes.

## Effect plots (ALEs)

## Convergence


# Simulating with FINN

## Assessing the model fit

## Simulate from bare ground



# Why FINN is different

FINN will be valid within the observed climatic envelop.







