---
title: "Fitting FINN to forest inventory data (Oregon, US FIA)"
format:
  html:
    toc: true
    toc-depth: 3
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{Fitting FINN to forest inventory data (Oregon, US FIA)}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Dynamic forest models are usually hand-calibrated. Here we calibrate FINN
directly from data: a subset of the US Forest Inventory & Analysis (FIA) program
for Oregon, prepared exactly as in the **Preparing your data for FINN** vignette.
Training is expensive, so it is *not* run when this page is knit — we show the fit
call, then load a pre-fitted model shipped with the package.

```{r}
#| label: libraries
library(FINN)
library(torch)
library(data.table)
library(ggplot2)
```

## The data

```{r}
#| label: load-data
ext <- function(f) system.file("extdata", f, package = "FINN")
obs_dt        <- fread(ext("fia_obs_dt.csv"))
env_dt        <- fread(ext("fia_env_dt.csv"))
env_scales_dt <- fread(ext("fia_env_scales_dt.csv"))
init_trees    <- fread(ext("fia_init_trees.csv"))
species_dt    <- fread(ext("fia_species_dt.csv"))

cat(sprintf("%d sites, %d species, observation years %s\n",
            uniqueN(obs_dt$siteID), uniqueN(obs_dt$species),
            paste(sort(unique(obs_dt$year)), collapse = " & ")))
species_dt
```

The environment is z-scaled; `env_scales_dt` holds the mean/sd so we can read it
back in natural units (°C, mm):

```{r}
#| label: plot-env
#| fig-width: 8
#| fig-height: 3
env_vars <- env_scales_dt$variable
env_nat  <- copy(env_dt)
for (v in env_vars) {
  s <- env_scales_dt[variable == v]
  env_nat[[v]] <- env_dt[[v]] * s$sd + s$mean
}
ggplot(melt(env_nat[, c("temp", "prec")], measure.vars = c("temp", "prec")),
       aes(value)) +
  geom_histogram(bins = 20, fill = "grey40") +
  facet_wrap(~variable, scales = "free", labeller = as_labeller(
    c(temp = "Mean~annual~temp~(degree*C)", prec = "Annual~precip~(mm)"), label_parsed)) +
  labs(x = NULL, y = "site x year") + theme_minimal()
```

A few observed series — basal area and tree numbers by year for the most
abundant species:

```{r}
#| label: plot-observations
#| fig-width: 8
#| fig-height: 3.5
top_sp <- obs_dt[, .(tot = sum(ba, na.rm = TRUE)), by = species][order(-tot)][1:4, species]
obs_long <- melt(obs_dt[species %in% top_sp],
                 id.vars = c("siteID", "year", "species", "species_name"),
                 measure.vars = c("ba", "trees"))
ggplot(obs_long, aes(factor(year), value, fill = factor(species))) +
  geom_boxplot(outlier.size = 0.4) +
  facet_wrap(~variable, scales = "free_y") +
  scale_fill_viridis_d(name = "species") +
  labs(x = "observation year", y = NULL) + theme_minimal()
```

## Initial cohorts

```{r}
#| label: init-cohorts
init_cohorts <- makeInitCohorts(init_trees, Nspecies = max(obs_dt$species))
```

## Build a Process-FINN

Every process is set `optimize* = TRUE`, so its species and environmental
parameters are learned from data. The `~.` formula lets every environmental
predictor enter the growth, regeneration and mortality responses.

```{r}
#| label: build-model
m <- finn(
  N_species            = uniqueN(obs_dt$species),
  recruits_dbh         = 12.9,
  competition_process  = createProcess(~0, FINN::competition,  optimizeSpecies = TRUE),
  growth_process       = createProcess(~., FINN::growth,       optimizeSpecies = TRUE, optimizeEnv = TRUE),
  regeneration_process = createProcess(~., FINN::regeneration, optimizeSpecies = TRUE, optimizeEnv = TRUE),
  mortality_process    = createProcess(~., FINN::mortality,    optimizeSpecies = TRUE, optimizeEnv = TRUE)
)
```

## Fit

Fitting runs gradient descent through the whole simulation. It is too slow for a
package build, so the chunk below is shown but not executed — it is the exact
call used to produce the shipped model (`dev/train_fia_model.R`):

```{r}
#| label: fit
#| eval: false
m$fit(
  env         = env_dt,
  data        = obs_dt,
  init_cohort = init_cohorts,
  device      = "cpu",
  epochs      = 3000L,
  batchsize   = 40L,
  patches     = 4,
  patch_size  = 0.06,
  weights     = c(0.1, 10, 1.0, 10.0, 1, 1),
  lr          = 0.01
)
torch::torch_save(m, "inst/extdata/fia_process_finn.pt")
```

```{r}
#| label: load-fitted
# Load the pre-fitted model shipped with the package and use it from here on.
m <- torch::torch_load(ext("fia_process_finn.pt"))
```

## Convergence

```{r}
#| label: convergence
#| fig-width: 6
#| fig-height: 3
loss <- data.table(epoch = seq_along(m$history),
                   loss = sapply(m$history, sum))
ggplot(loss, aes(epoch, loss)) +
  geom_line(colour = "steelblue") +
  scale_y_log10() +
  labs(x = "epoch", y = "total training loss (log scale)") + theme_minimal()
```

## Assess the fit

We simulate from the fitted model and compare predictions to observations,
per response variable (Spearman correlation, matched on site × year × species):

```{r}
#| label: assess
sim  <- m$simulate(env = env_dt, init_cohort = init_cohorts,
                   patches = 4, patch_size = 0.06, device = "cpu")
pred <- sim$long$site[, .(pred = mean(value)), by = .(siteID, year, species, variable)]
pred[, variable := as.character(variable)]

obs_m <- melt(obs_dt, id.vars = c("siteID", "year", "species", "species_name"),
              variable.name = "variable", value.name = "obs")
obs_m[, variable := as.character(variable)]

cmp <- merge(obs_m, pred, by = c("siteID", "year", "species", "variable"))
fit_metrics <- cmp[!is.na(obs) & !is.na(pred),
                   .(n = .N, spearman = round(cor(obs, pred, method = "spearman"), 2)),
                   by = variable][order(-spearman)]
fit_metrics
```

The structural variables — basal area, tree numbers and diameter — are
reconstructed well; the noisier per-tree rates (growth, mortality, regeneration)
are harder to constrain from two inventories.

```{r}
#| label: assess-plot
#| fig-width: 7
#| fig-height: 3
ggplot(cmp[variable %in% c("ba", "trees", "dbh") & !is.na(obs)],
       aes(obs, pred)) +
  geom_point(alpha = 0.25, size = 0.6) +
  geom_abline(slope = 1, intercept = 0, colour = "red", linewidth = 0.3) +
  facet_wrap(~variable, scales = "free") +
  labs(x = "observed", y = "predicted") + theme_minimal()
```

## Inspect the niches

Because the environmental responses are learned, we can read off each species'
**niche** — how its growth responds to climate. We evaluate the fitted growth
environmental response over each climate gradient (varying one variable across
its observed range, holding the others at their mean) and back-transform to
natural units. (FINN also ships `ALE()` for accumulated-local-effect curves; the
direct response below is clearer on this small sample.)

```{r}
#| label: niches
#| fig-width: 8.5
#| fig-height: 3.6
env_vars <- env_scales_dt$variable

niche <- rbindlist(lapply(c("temp", "prec"), function(target) {
  rng  <- range(env_dt[[target]])                 # observed z-range, no extrapolation
  grid <- seq(rng[1], rng[2], length.out = 60)
  X <- matrix(0, length(grid), length(env_vars) + 1); X[, 1] <- 1   # intercept
  X[, which(env_vars == target) + 1] <- grid
  eff <- as.matrix(m$nn_growth(torch_tensor(X, dtype = torch_float32())))
  s <- env_scales_dt[variable == target]
  rbindlist(lapply(seq_len(ncol(eff)), function(j)
    data.table(var = target, species = j,
               x = grid * s$sd + s$mean, growth_mult = exp(eff[, j]))))
}))
niche <- merge(niche, species_dt, by = "species")

ggplot(niche, aes(x, growth_mult, colour = species_name)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~var, scales = "free", labeller = as_labeller(
    c(temp = "temperature~(degree*C)", prec = "precipitation~(mm)"), label_parsed)) +
  scale_colour_viridis_d(name = "species") +
  labs(x = NULL, y = "growth response (env. multiplier)") + theme_minimal()
```

Each curve is an inferred environmental response: how a species' growth scales
along the temperature and precipitation gradients, learned purely from the
inventory data. Ecologically familiar contrasts emerge — e.g. *Pinus ponderosa*
grows best on the driest sites while wetter-climate species ramp up with
precipitation. This is the bridge to using FINN for niche inference and
projection.

```{r}
#| label: session
sessionInfo()
```
