---
title: "Fitting FINN to forest inventory data (Oregon, US FIA)"
format:
  html:
    toc: true
    toc-depth: 3
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{Fitting FINN to forest inventory data (Oregon, US FIA)}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Dynamic forest models are usually hand-calibrated. Here we calibrate FINN
directly from data: a subset of the US Forest Inventory & Analysis (FIA) program
for Oregon, prepared exactly as in the **Preparing your data for FINN** vignette.
Training is expensive, so it is *not* run when this page is knit — we show the fit
call, then load a pre-fitted model shipped with the package.

```{r}
#| label: libraries
library(FINN)
library(torch)
library(data.table)
library(ggplot2)
```

## The data

```{r}
#| label: load-data
ext <- function(f) system.file("extdata", f, package = "FINN")
obs_dt     <- fread(ext("fia_obs_dt.csv"))
env_dt     <- fread(ext("fia_env_dt.csv"))     # RAW, untransformed climate
init_trees <- fread(ext("fia_init_trees.csv"))
species_dt <- fread(ext("fia_species_dt.csv"))

cat(sprintf("%d sites, %d species, observation years %s\n",
            uniqueN(obs_dt$siteID), uniqueN(obs_dt$species),
            paste(sort(unique(obs_dt$year)), collapse = " & ")))
species_dt
```

The environment is supplied in **natural units** — FINN z-standardizes the
predictors internally during fitting (`env_autoscale = TRUE`, the default) and reuses
the same constants at prediction time, so we never have to scale by hand:

```{r}
#| label: plot-env
#| fig-width: 8
#| fig-height: 3
ggplot(melt(env_dt[, c("temp", "prec")], measure.vars = c("temp", "prec")),
       aes(value)) +
  geom_histogram(bins = 20, fill = "grey40") +
  facet_wrap(~variable, scales = "free", labeller = as_labeller(
    c(temp = "Mean~annual~temp~(degree*C)", prec = "Annual~precip~(mm)"), label_parsed)) +
  labs(x = NULL, y = "site x year") + theme_minimal()
```

A few observed series — basal area and tree numbers by year for the most
abundant species:

```{r}
#| label: plot-observations
#| fig-width: 8
#| fig-height: 3.5
top_sp <- obs_dt[, .(tot = sum(ba, na.rm = TRUE)), by = species][order(-tot)][1:4, species]
obs_long <- melt(obs_dt[species %in% top_sp],
                 id.vars = c("siteID", "year", "species", "species_name"),
                 measure.vars = c("ba", "trees"))
ggplot(obs_long, aes(factor(year), value, fill = factor(species))) +
  geom_boxplot(outlier.size = 0.4) +
  facet_wrap(~variable, scales = "free_y") +
  scale_fill_viridis_d(name = "species") +
  labs(x = "observation year", y = NULL) + theme_minimal()
```

## Initial cohorts

```{r}
#| label: init-cohorts
init_cohorts <- makeInitCohorts(init_trees, Nspecies = max(obs_dt$species))
```

## Build a Process-FINN

Every process is set `optimize* = TRUE`, so its species and environmental
parameters are learned from data. The `~.` formula lets every environmental
predictor enter the growth, regeneration and mortality responses.

```{r}
#| label: build-model
m <- finn(
  N_species            = uniqueN(obs_dt$species),
  recruits_dbh         = 12.9,
  competition_process  = createProcess(~0, FINN::competition,  optimizeSpecies = TRUE),
  growth_process       = createProcess(~., FINN::growth,       optimizeSpecies = TRUE, optimizeEnv = TRUE),
  regeneration_process = createProcess(~., FINN::regeneration, optimizeSpecies = TRUE, optimizeEnv = TRUE),
  mortality_process    = createProcess(~., FINN::mortality,    optimizeSpecies = TRUE, optimizeEnv = TRUE)
)
```

## Fit

Fitting runs gradient descent through the whole simulation. It is too slow for a
package build, so the chunk below is shown but not executed — it is the exact
call used to produce the shipped model (`dev/train_fia_model.R`):

```{r}
#| label: fit
#| eval: false
m$fit(
  env         = env_dt,      # raw climate — FINN standardizes it internally
  data        = obs_dt,
  init_cohort = init_cohorts,
  device      = "cpu",
  epochs      = 3000L,
  batchsize   = 40L,
  patches     = 4,
  patch_size  = 0.06,
  weights     = c(0.1, 10, 1.0, 10.0, 1, 1),
  lr          = 0.01,
  env_autoscale   = TRUE         # default; learns & stores the predictor scaling
)
torch::torch_save(m, "inst/extdata/fia_process_finn.pt")
```

```{r}
#| label: load-fitted
# Load the pre-fitted model shipped with the package and use it from here on.
m <- torch::torch_load(ext("fia_process_finn.pt"))
```

With `env_autoscale = TRUE`, the per-predictor mean and standard deviation learned
from the training climate are stored on the model and re-applied automatically
whenever we predict — so raw `env_dt` is all we ever pass:

```{r}
#| label: env-scaling
m$env_scaling
```

## Convergence

```{r}
#| label: convergence
#| fig-width: 6
#| fig-height: 3
loss <- data.table(epoch = seq_along(m$history),
                   loss = sapply(m$history, sum))
ggplot(loss, aes(epoch, loss)) +
  geom_line(colour = "steelblue") +
  scale_y_log10() +
  labs(x = "epoch", y = "total training loss (log scale)") + theme_minimal()
```

## Assess the fit

We simulate from the fitted model and compare predictions to observations,
per response variable (Spearman correlation, matched on site × year × species):

```{r}
#| label: assess
sim  <- m$simulate(env = env_dt, init_cohort = init_cohorts,
                   patches = 4, patch_size = 0.06, device = "cpu")
pred <- sim$long$site[, .(pred = mean(value)), by = .(siteID, year, species, variable)]
pred[, variable := as.character(variable)]

obs_m <- melt(obs_dt, id.vars = c("siteID", "year", "species", "species_name"),
              variable.name = "variable", value.name = "obs")
obs_m[, variable := as.character(variable)]

cmp <- merge(obs_m, pred, by = c("siteID", "year", "species", "variable"))
fit_metrics <- cmp[!is.na(obs) & !is.na(pred),
                   .(n = .N, spearman = round(cor(obs, pred, method = "spearman"), 2)),
                   by = variable][order(-spearman)]
fit_metrics
```

The structural variables — basal area, tree numbers and diameter — are
reconstructed well; the noisier per-tree rates (growth, mortality, regeneration)
are harder to constrain from two inventories.

```{r}
#| label: assess-plot
#| fig-width: 7
#| fig-height: 3
ggplot(cmp[variable %in% c("ba", "trees", "dbh") & !is.na(obs)],
       aes(obs, pred)) +
  geom_point(alpha = 0.25, size = 0.6) +
  geom_abline(slope = 1, intercept = 0, colour = "red", linewidth = 0.3) +
  facet_wrap(~variable, scales = "free") +
  labs(x = "observed", y = "predicted") + theme_minimal()
```

## Inspect the niches

Because the environmental responses are learned, we can read off each species'
**niche** — how its growth responds to climate. We vary one climate variable
across its observed (natural-unit) range, hold the others at their mean, and
evaluate the fitted growth response. Following the convention of interpretable-ML
tools (e.g. cito's effect plots), we plot the response on the **natural predictor
axis** — using the model's own stored standardization (`m$env_scaling`) to map the
natural grid onto the scale the network was trained on.

```{r}
#| label: niches
#| fig-width: 8.5
#| fig-height: 3.6
scaling  <- as.data.table(m$env_scaling)
env_vars <- scaling$variable

niche <- rbindlist(lapply(c("temp", "prec"), function(target) {
  s    <- scaling[variable == target]
  grid <- seq(min(env_dt[[target]]), max(env_dt[[target]]), length.out = 60)  # natural units
  X <- matrix(0, length(grid), length(env_vars) + 1); X[, 1] <- 1   # intercept; others at mean (=0 scaled)
  X[, which(env_vars == target) + 1] <- (grid - s$center) / s$scale # standardize the varied predictor
  eff <- as.matrix(m$nn_growth(torch_tensor(X, dtype = torch_float32())))
  rbindlist(lapply(seq_len(ncol(eff)), function(j)
    data.table(var = target, species = j, x = grid, growth_mult = exp(eff[, j]))))
}))
niche <- merge(niche, species_dt, by = "species")

ggplot(niche, aes(x, growth_mult, colour = species_name)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~var, scales = "free", labeller = as_labeller(
    c(temp = "temperature~(degree*C)", prec = "precipitation~(mm)"), label_parsed)) +
  scale_colour_viridis_d(name = "species") +
  labs(x = NULL, y = "growth response (env. multiplier)") + theme_minimal()
```

Each curve is an inferred environmental response: how a species' growth scales
along the temperature and precipitation gradients, learned purely from the
inventory data. Ecologically familiar contrasts emerge — e.g. *Pinus ponderosa*
grows best on the driest sites while wetter-climate species ramp up with
precipitation. This is the bridge to using FINN for niche inference and
projection.

## Replacing a process with a neural network (Hybrid-FINN)

FINN's defining feature is that **any single demographic process can be swapped
from a mechanistic function to a neural network**, while the others stay
mechanistic — a *hybrid* model. The mechanistic processes keep the system
ecologically constrained; the network absorbs structure the fixed functional form
cannot express. Here we replace **growth** with a small neural network via
`createHybrid()`, leaving competition, regeneration and mortality mechanistic, and
refit on the same data.

```{r}
#| label: hybrid-build
m_hybrid <- finn(
  N_species            = uniqueN(obs_dt$species),
  recruits_dbh         = 12.9,
  competition_process  = createProcess(~0, FINN::competition,  optimizeSpecies = TRUE),
  growth_process       = createHybrid(~., hidden = c(20L, 20L), transformer = FALSE),  # NN replaces growth
  regeneration_process = createProcess(~., FINN::regeneration, optimizeSpecies = TRUE, optimizeEnv = TRUE),
  mortality_process    = createProcess(~., FINN::mortality,    optimizeSpecies = TRUE, optimizeEnv = TRUE)
)
```

Fitting is identical — the same `fit()` call (and `env_autoscale = TRUE`) — and is
again run by `dev/train_fia_model.R`, so the chunk below is shown but not executed:

```{r}
#| label: hybrid-fit
#| eval: false
m_hybrid$fit(
  env = env_dt, data = obs_dt, init_cohort = init_cohorts, device = "cpu",
  epochs = 3000L, batchsize = 40L, patches = 4, patch_size = 0.06,
  weights = c(0.1, 10, 1.0, 10.0, 1, 1), lr = 0.01, env_autoscale = TRUE
)
torch::torch_save(m_hybrid, "inst/extdata/fia_hybrid_finn.pt")
```

```{r}
#| label: hybrid-load
m_hybrid <- torch::torch_load(ext("fia_hybrid_finn.pt"))
```

How does the learned-growth hybrid compare to the fully mechanistic model? We
score both the same way — predicted-vs-observed **Spearman correlation** (rank
agreement) and **RMSE** (in each response's own units) per variable:

```{r}
#| label: hybrid-compare
metrics_of <- function(model) {
  model$eval()                                   # deterministic: dropout off
  sim <- model$simulate(env = env_dt, init_cohort = init_cohorts,
                        patches = 4, patch_size = 0.06, device = "cpu")
  p <- sim$long$site[, .(pred = mean(value)), by = .(siteID, year, species, variable)]
  p[, variable := as.character(variable)]
  cc <- merge(obs_m, p, by = c("siteID", "year", "species", "variable"))
  cc[!is.na(obs) & !is.na(pred),
     .(spearman = round(cor(obs, pred, method = "spearman"), 2),
       rmse     = round(sqrt(mean((obs - pred)^2)), 2)), by = variable]
}
comparison <- merge(
  metrics_of(m)[,        .(variable, spearman_proc = spearman, rmse_proc = rmse)],
  metrics_of(m_hybrid)[, .(variable, spearman_hyb  = spearman, rmse_hyb  = rmse)],
  by = "variable")[order(-spearman_proc)]
comparison
```

The hybrid reproduces the structural variables about as well as the mechanistic
model — with no growth equation specified. Hybrids are the tool of choice when a
process's mechanistic form is uncertain or you suspect the data carry structure
the equation misses; the surrounding mechanistic processes still anchor the model
ecologically.

### Did the network recover the mechanistic niche?

The central comparison in the FINN paper is whether the neural network learned the
*same* environmental response as the mechanistic growth function. We compare them
with **accumulated local effects (ALE)** — the effect measure of choice here
because the bioclimatic predictors are correlated, where partial-dependence curves
would extrapolate to unrealistic climate combinations. ALE instead accumulates
*local* changes in predicted growth within the observed climate distribution.

For the hybrid we also draw a **confidence band**, exactly as in the FINNetAl
study: the growth network is run with **dropout left on** (Monte-Carlo dropout)
over bootstrap resamples, and we summarise the resulting ensemble of ALE curves as
mean ± 1 SD. The mechanistic model is deterministic, so it is a single line.

ALE is centred at zero by construction; to keep the y-axis interpretable we add
each species' **mean predicted growth** back, so the curves are on the absolute
growth scale (cm yr⁻¹, ≥ 0). We compare the two models across **four drivers of
growth** — diameter and available light (stand state), and temperature and
precipitation (climate) — for four representative species. The driver values are
the ones the cohorts actually experience in a simulation (recovered through the
model's `record_raws` mechanism), and the bins are taken from **each species' own**
realized range, so every curve stays strictly within the data that species
occupies — **no extrapolation**, as in FINNetAl.

```{r}
#| label: niche-compare
#| fig-width: 9.5
#| fig-height: 9
env_vars <- as.data.table(m$env_scaling)$variable
VARS <- c("dbh", "light", "temp", "prec"); SPP <- 1:4

# realized cohort states (dbh, light, trees + climate) from a simulation
get_states <- function(model) {
  model$eval(); model$record_raws <- TRUE
  invisible(model$simulate(env = env_dt, init_cohort = init_cohorts,
                           patches = 4, patch_size = 0.06, device = "cpu"))
  rg <- model$raw_g; model$record_raws <- FALSE; model$raw_g <- NULL
  env_cols <- c("siteID", env_vars)
  env_site <- unique(env_dt, by = "siteID")[, ..env_cols]          # one climate row per site
  st <- rbindlist(lapply(seq_along(rg), function(t) {
    a <- rg[[t]]; d <- dim(a)
    g <- as.data.table(expand.grid(site = 1:d[1], patch = 1:d[2], cohort = 1:d[3]))
    g[, `:=`(dbh = as.vector(a[,,,1]), light = as.vector(a[,,,2]),
             trees = as.vector(a[,,,3]), species = as.vector(a[,,,4]))]
    merge(g, env_site, by.x = "site", by.y = "siteID")
  }))
  st[trees > 0.5 & dbh > 0.5]
}

# predict growth for a table of states (climate standardized internally)
make_predf <- function(model) {
  scaling <- as.data.table(model$env_scaling); is_h <- inherits(model$process_growth, "hybrid")
  function(st, sp) {
    n <- nrow(st)
    Z <- do.call(cbind, lapply(env_vars, function(v){ s<-scaling[variable==v]; (st[[v]]-s$center)/s$scale }))
    mk <- function(v, ty) torch_tensor(array(v, c(n,1,1)), dtype = ty)
    pred <- torch_tensor(cbind(1, Z), dtype = torch_float32()); species <- mk(sp, torch_long())
    if (!is_h) { pred <- model$nn_growth(pred); pred <- FINN::index_species(pred, species) }
    as.numeric(as_array(model$growth_func(dbh = mk(st$dbh, torch_float32()),
      trees = mk(st$trees, torch_float32()), light = mk(st$light, torch_float32()),
      species = species, pred = pred, parGrowth = model$par_growth)))
  }
}

# ALE of one driver for one species (raw units; anchored + floored at 0)
ale_feature <- function(predf, st_sp, sp, target, breaks) {
  K <- length(breaks) - 1L
  bin <- cut(st_sp[[target]], breaks, include.lowest = TRUE, labels = FALSE)
  local <- numeric(K)
  for (k in seq_len(K)) {
    idx <- which(bin == k); if (!length(idx)) next
    lo <- copy(st_sp[idx]); hi <- copy(st_sp[idx])
    set(lo, j = target, value = breaks[k]); set(hi, j = target, value = breaks[k+1])
    local[k] <- mean(predf(hi, sp) - predf(lo, sp))
  }
  ale <- cumsum(local) - mean(cumsum(local)) + mean(predf(st_sp, sp))
  data.table(x = (breaks[-1] + breaks[-(K+1)]) / 2, growth = pmax(ale, 0))
}

set.seed(1); FINN.seed(1)          # reproducible simulation states + MC-dropout
st_p <- get_states(m); st_h <- get_states(m_hybrid)

one_model <- function(model, st, mc, reps) {
  predf <- make_predf(model)
  rbindlist(lapply(VARS, function(v) rbindlist(lapply(SPP, function(sp) {
    st_sp <- st[species == sp]
    if (nrow(st_sp) < 30) return(NULL)
    # trim the sparse top 5% of this driver, then bin from the species' own
    # range -> the ALE stays within well-sampled data (no extrapolation)
    st_sp  <- st_sp[get(v) <= quantile(st_sp[[v]], 0.95)]
    breaks <- unique(quantile(st_sp[[v]], probs = seq(0, 1, length.out = 9)))
    if (length(breaks) < 3) return(NULL)
    curves <- rbindlist(lapply(seq_len(reps), function(j) {
      if (mc) model$train() else model$eval()                     # MC-dropout on / off
      d <- if (mc) st_sp[sample(.N, replace = TRUE)] else st_sp    # bootstrap
      ale_feature(predf, d, sp, v, breaks)
    }))
    curves[, .(growth = mean(growth), lo = pmax(mean(growth)-sd(growth),0),
               hi = mean(growth)+sd(growth)), by = x][, `:=`(variable = v, species = sp)]
  }))))
}

grid <- rbind(
  one_model(m_hybrid, st_h, mc = TRUE,  reps = 15)[, model := "Hybrid (growth = NN)"],
  one_model(m,        st_p, mc = FALSE, reps = 1 )[, model := "Process (mechanistic)"])
grid <- merge(grid, species_dt, by = "species")

library(patchwork)
cols <- c("Process (mechanistic)" = "#1b9e77", "Hybrid (growth = NN)" = "#d95f02")
# one row per driver; facet_wrap over species shares the x-axis across species
# (scales = "free_y"), and the driver name becomes the x-axis title below the row.
make_row <- function(v, xlab, top) {
  g <- ggplot(grid[variable == v], aes(x, growth, colour = model, fill = model)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.18, colour = NA) +
    geom_line(linewidth = 0.6) +
    facet_wrap(~species_name, nrow = 1, scales = "free_y") +
    scale_colour_manual(values = cols) + scale_fill_manual(values = cols) +
    labs(x = xlab, y = "growth (cm/yr)", colour = NULL, fill = NULL) +
    theme_minimal() + theme(legend.position = "none")
  if (!top) g <- g + theme(strip.text = element_blank())
  g
}
(make_row("dbh",  "dbh (cm)", TRUE) /
 make_row("light","available light", FALSE) /
 make_row("temp", expression(temperature ~ (degree * C)), FALSE) /
 make_row("prec", "precipitation (mm)", FALSE)) +
  plot_layout(guides = "collect") & theme(legend.position = "top")
```

Each panel asks whether the network reproduced the mechanistic response to one
driver. Where the green line stays inside the hybrid's band, the network recovered
the mechanistic relationship from data alone; where they part — often at the
climate extremes or for the more sparsely observed species — the network settled
on a different, data-driven shape. The companion article *Recovering a forest
process with Hybrid-FINN* makes the same comparison against known simulated truth.

```{r}
#| label: session
sessionInfo()
```
