---
title: "Fitting FINN to forest inventory data (Oregon, US FIA)"
format:
  html:
    toc: true
    toc-depth: 3
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{Fitting FINN to forest inventory data (Oregon, US FIA)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Dynamic forest models are usually hand-calibrated. Here we calibrate
FINN directly from data: a subset of the US Forest Inventory & Analysis
(FIA) program for Oregon, prepared exactly as in the **Preparing your
data for FINN** vignette. Training — and even loading the fitted model — needs a torch backend, so
the torch steps below are shown but **not executed** when this page is
knit; their results are precomputed (`dev/precompute_vignettes.R`) and
loaded from a small cache, so the vignette builds anywhere.

```{r}
#| label: libraries
library(FINN)
library(torch)
library(data.table)
library(ggplot2)
```

## The data

These four bundled tables are built by **`dev/make_extdata.R`**, which subsamples
the full Oregon FIA set in `data-raw/` down to ~40 sites. The **climate**
(`fia_env_dt.csv`) traces upstream to the FINN-fia analysis repo
(`scripts/03_attach_environment.R` → `07_prepare_finn_inputs.R`); see
`data-raw/README.md` for the full chain.

```{r}
#| label: load-data
ext <- function(f) system.file("extdata", f, package = "FINN")
obs_dt     <- fread(ext("fia_obs_dt.csv"))
env_dt     <- fread(ext("fia_env_dt.csv"))     # RAW, untransformed climate
init_trees <- fread(ext("fia_init_trees.csv"))
species_dt <- fread(ext("fia_species_dt.csv"))

cat(sprintf("%d sites, %d species, observation years %s\n",
            uniqueN(obs_dt$siteID), uniqueN(obs_dt$species),
            paste(sort(unique(obs_dt$year)), collapse = " & ")))
species_dt
```

The environment is supplied in **natural units**: with
`env_autoscale = TRUE` (the default) FINN z-standardizes the predictors
internally and reuses the same constants at prediction time, so raw
values are all we ever pass.

```{r}
#| label: plot-env
#| fig-width: 8
#| fig-height: 3
ggplot(melt(env_dt[, c("temp", "prec")], measure.vars = c("temp", "prec")),
       aes(value)) +
  geom_histogram(bins = 20, fill = "grey40") +
  facet_wrap(~variable, scales = "free", labeller = as_labeller(
    c(temp = "Mean~annual~temp~(degree*C)", prec = "Annual~precip~(mm)"), label_parsed)) +
  labs(x = NULL, y = "site x year") + theme_minimal()
```

A few observed series — basal area and tree numbers by year for the most
abundant species:

```{r}
#| label: plot-observations
#| fig-width: 8
#| fig-height: 3.5

top_sp <- obs_dt[species_name != "other", .(tot = sum(ba, na.rm = TRUE)), by = species][order(-tot)][1:4, species]
obs_long <- melt(obs_dt[species %in% top_sp],
                 id.vars = c("siteID", "year", "species", "species_name"),
                 measure.vars = c("ba", "trees"))
ggplot(obs_long, aes(factor(year), value, fill = factor(species_name))) +
  geom_boxplot(outlier.size = 0.4) +
  facet_wrap(~variable, scales = "free_y") +
  scale_fill_viridis_d(name = "species") +
  labs(x = "observation year", y = NULL) + theme_minimal()
```

## Initial cohorts

```{r}
#| label: init-cohorts
#| eval: false
init_cohorts <- makeInitCohorts(init_trees, Nspecies = max(obs_dt$species))
```

```{r}
#| label: precomputed
#| include: false
# Torch results, precomputed by dev/precompute_vignettes.R and shipped with the
# package so this vignette renders without a torch backend.
.vig <- readRDS(system.file("extdata", "vig_fit_to_fia.rds", package = "FINN"))
loss <- .vig$loss; fit_metrics <- .vig$fit_metrics; cmp <- .vig$cmp
niche <- .vig$niche; comparison <- .vig$comparison; growth <- .vig$growth
```

## Build a Process-FINN

Every process is set `optimize* = TRUE`, so its species and
environmental parameters are learned from data. The `~.` formula lets
every environmental predictor enter the growth, regeneration and
mortality responses.

```{r}
#| label: build-model
#| eval: false
m <- finn(
  N_species            = uniqueN(obs_dt$species),
  recruits_dbh         = 12.9,
  competition_process  = createProcess(~0, FINN::competition,  optimizeSpecies = TRUE),
  growth_process       = createProcess(~., FINN::growth,       optimizeSpecies = TRUE, optimizeEnv = TRUE),
  regeneration_process = createProcess(~., FINN::regeneration, optimizeSpecies = TRUE, optimizeEnv = TRUE),
  mortality_process    = createProcess(~., FINN::mortality,    optimizeSpecies = TRUE, optimizeEnv = TRUE)
)
```

## Fit

Fitting runs gradient descent through the whole simulation. It is too
slow for a package build, so the chunk below is shown but not executed —
it is the exact call used to produce the shipped model
(`dev/train_fia_model.R`):

```{r}
#| label: fit
#| eval: false
m$fit(
  env         = env_dt,      # raw climate
  data        = obs_dt,
  init_cohort = init_cohorts,
  device      = "cpu",
  epochs      = 3000L,
  batchsize   = 40L,
  patches     = 4,
  patch_size  = 0.06,
  weights     = c(0.1, 10, 1.0, 10.0, 1, 1),
  lr          = 0.01,
  env_autoscale   = TRUE         # default
)
torch::torch_save(m, "inst/extdata/fia_process_finn.pt")
```

```{r}
#| label: load-fitted
#| eval: false
# Load the pre-fitted model shipped with the package and use it from here on.
m <- torch::torch_load(ext("fia_process_finn.pt"))
```

## Convergence

```{r}
#| label: convergence-code
#| eval: false
loss <- data.table(epoch = seq_along(m$history),
                   loss = sapply(m$history, sum))
```

```{r}
#| label: convergence-plot
#| echo: false
#| fig-width: 6
#| fig-height: 3
ggplot(loss, aes(epoch, loss)) +
  geom_line(colour = "steelblue") +
  scale_y_log10() +
  labs(x = "epoch", y = "total training loss (log scale)") + theme_minimal()
```

## Assess the fit

We simulate from the fitted model and compare predictions to
observations, per response variable (Spearman correlation, matched on
site × year × species):

```{r}
#| label: assess
#| eval: false
sim  <- m$simulate(env = env_dt, init_cohort = init_cohorts,
                   patches = 4, patch_size = 0.06, device = "cpu")
pred <- sim$long$site[, .(pred = mean(value)), by = .(siteID, year, species, variable)]
pred[, variable := as.character(variable)]

obs_m <- melt(obs_dt, id.vars = c("siteID", "year", "species", "species_name"),
              variable.name = "variable", value.name = "obs")
obs_m[, variable := as.character(variable)]

cmp <- merge(obs_m, pred, by = c("siteID", "year", "species", "variable"))
fit_metrics <- cmp[!is.na(obs) & !is.na(pred),
                   .(n = .N, spearman = round(cor(obs, pred, method = "spearman"), 2)),
                   by = variable][order(-spearman)]
fit_metrics
```

```{r}
#| label: assess-metrics
#| echo: false
fit_metrics
```

The structural variables — basal area, tree numbers and diameter — are
reconstructed well; the noisier per-tree rates (growth, mortality,
regeneration) are harder to constrain from two inventories.

```{r}
#| label: assess-plot
#| fig-width: 7
#| fig-height: 3
ggplot(cmp[variable %in% c("ba", "trees", "dbh") & !is.na(obs)],
       aes(obs, pred)) +
  geom_point(alpha = 0.25, size = 0.6) +
  geom_abline(slope = 1, intercept = 0, colour = "red", linewidth = 0.3) +
  facet_wrap(~variable, scales = "free") +
  labs(x = "observed", y = "predicted") + theme_minimal()
```

## Inspect the niches

Because the environmental responses are learned, we can read off each
species' **niche** — how its growth responds to climate. We vary one
climate variable across its observed (natural-unit) range, hold the
others at their mean, and evaluate the fitted growth response on the
**natural predictor axis**, as in cito's effect plots.

```{r}
#| label: niches-code
#| eval: false
scaling  <- as.data.table(m$env_scaling)
env_vars <- scaling$variable

niche <- rbindlist(lapply(c("temp", "prec"), function(target) {
  s    <- scaling[variable == target]
  grid <- seq(min(env_dt[[target]]), max(env_dt[[target]]), length.out = 60)  # natural units
  X <- matrix(0, length(grid), length(env_vars) + 1); X[, 1] <- 1   # intercept; others at mean (=0 scaled)
  X[, which(env_vars == target) + 1] <- (grid - s$center) / s$scale # standardize the varied predictor
  eff <- as.matrix(m$nn_growth(torch_tensor(X, dtype = torch_float32())))
  rbindlist(lapply(seq_len(ncol(eff)), function(j)
    data.table(var = target, species = j, x = grid, growth_mult = exp(eff[, j]))))
}))
niche <- merge(niche, species_dt, by = "species")
```

```{r}
#| label: niches-plot
#| fig-width: 8.5
#| fig-height: 3.6
ggplot(niche, aes(x, growth_mult, colour = species_name)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~var, scales = "free", labeller = as_labeller(
    c(temp = "temperature~(degree*C)", prec = "precipitation~(mm)"), label_parsed)) +
  scale_colour_viridis_d(name = "species") +
  labs(x = NULL, y = "growth response (env. multiplier)") + theme_minimal()
```

Each curve is an inferred environmental response: how a species' growth
scales along the temperature and precipitation gradients, learned purely
from the inventory data. Ecologically familiar contrasts emerge — e.g.
*Pinus ponderosa* grows best on the driest sites while wetter-climate
species ramp up with precipitation. This is the bridge to using FINN for
niche inference and projection.

## Replacing a process with a neural network (Hybrid-FINN)

FINN's defining feature is that **any single demographic process can be
swapped from a mechanistic function to a neural network**, while the
others stay mechanistic — a *hybrid* model. The mechanistic processes
keep the system ecologically constrained; the network absorbs structure
the fixed functional form cannot express. Here we replace **growth**
with a small neural network via `createHybrid()`, leaving competition,
regeneration and mortality mechanistic, and refit on the same data.

```{r}
#| label: hybrid-build
#| eval: false
m_hybrid <- finn(
  N_species            = uniqueN(obs_dt$species),
  recruits_dbh         = 12.9,
  competition_process  = createProcess(~0, FINN::competition,  optimizeSpecies = TRUE),
  growth_process       = createHybrid(~., hidden = c(20L, 20L), transformer = FALSE),  # NN replaces growth
  regeneration_process = createProcess(~., FINN::regeneration, optimizeSpecies = TRUE, optimizeEnv = TRUE),
  mortality_process    = createProcess(~., FINN::mortality,    optimizeSpecies = TRUE, optimizeEnv = TRUE)
)
```

Fitting is identical — the same `fit()` call — and is again run by
`dev/train_fia_model.R`, so the chunk below is shown but not executed:

```{r}
#| label: hybrid-fit
#| eval: false
m_hybrid$fit(
  env = env_dt, data = obs_dt, init_cohort = init_cohorts, device = "cpu",
  epochs = 3000L, batchsize = 40L, patches = 4, patch_size = 0.06,
  weights = c(0.1, 10, 1.0, 10.0, 1, 1), lr = 0.01, env_autoscale = TRUE
)
torch::torch_save(m_hybrid, "inst/extdata/fia_hybrid_finn.pt")
```

```{r}
#| label: hybrid-load
#| eval: false
m_hybrid <- torch::torch_load(ext("fia_hybrid_finn.pt"))
```

How does the learned-growth hybrid compare to the fully mechanistic
model? We score both the same way — predicted-vs-observed **Spearman
correlation** (rank agreement) and **RMSE** (in each response's own
units) per variable:

```{r}
#| label: hybrid-compare
#| eval: false
metrics_of <- function(model) {
  model$eval()                                   # deterministic: dropout off
  sim <- model$simulate(env = env_dt, init_cohort = init_cohorts,
                        patches = 4, patch_size = 0.06, device = "cpu")
  p <- sim$long$site[, .(pred = mean(value)), by = .(siteID, year, species, variable)]
  p[, variable := as.character(variable)]
  cc <- merge(obs_m, p, by = c("siteID", "year", "species", "variable"))
  cc[!is.na(obs) & !is.na(pred),
     .(spearman = round(cor(obs, pred, method = "spearman"), 2),
       rmse     = round(sqrt(mean((obs - pred)^2)), 2)), by = variable]
}
comparison <- merge(
  metrics_of(m)[,        .(variable, spearman_proc = spearman, rmse_proc = rmse)],
  metrics_of(m_hybrid)[, .(variable, spearman_hyb  = spearman, rmse_hyb  = rmse)],
  by = "variable")[order(-spearman_proc)]
comparison[,.(variable,spearman_proc,spearman_hyb,rmse_proc,rmse_hyb)]
```

```{r}
#| label: hybrid-compare-table
#| echo: false
comparison[, .(variable, spearman_proc, spearman_hyb, rmse_proc, rmse_hyb)]
```

The hybrid reproduces the structural variables about as well as the
mechanistic model — with no growth equation specified. Hybrids are the
tool of choice when a process's mechanistic form is uncertain or you
suspect the data carry structure the equation misses; the surrounding
mechanistic processes still anchor the model ecologically.

### Did the network recover the mechanistic response?

The point of a hybrid is to ask which functional form the neural network learned and how it *compares* to the response as the mechanistic function it replaced. FINN answers
this directly with `ALE()`, which simulates the model and then computes
**accumulated local effects** — the effect measure of choice when
predictors are correlated, because it accumulates *local* changes within
the observed data instead of extrapolating to unrealistic combinations.
It returns one table per process (`growth`, `mortality`,
`regeneration`), each giving the effect of every driver on that process.
Below we pull the growth tables and compare the response to tree size.

```{r}
#| label: niche-compare-code
#| eval: false
# ALE() simulates the model, then accumulates local effects for every
# process and driver (raw env_dt, exactly as for fit()).
FINN.seed(1)
ale_proc <- ALE(m,        env_dt, init_cohorts)   # mechanistic growth
ale_hyb  <- ALE(m_hybrid, env_dt, init_cohorts)   # growth replaced by an NN

growth <- rbind(
  data.table(ale_proc$growth)[, model := "Process (mechanistic)"],
  data.table(ale_hyb$growth)[,  model := "Hybrid (growth = NN)"])
```

```{r}
#| label: niche-compare-plot
#| fig-width: 9.5
#| fig-height: 3.2
# growth response to tree size, mechanistic vs hybrid, for four well-sampled species
growth_dbh <- merge(growth[var == "dbh" & species %in% c(5, 2, 1, 4)],
                species_dt, by = "species")

ggplot(growth_dbh, aes(x, ale, colour = model)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~species_name, nrow = 1, scales = "free") +
  scale_colour_manual(values = c("Process (mechanistic)" = "#1b9e77",
                                 "Hybrid (growth = NN)"  = "#d95f02")) +
  labs(x = "diameter (cm)", y = "accumulated local effect on growth",
       colour = NULL) +
  theme_minimal() + theme(legend.position = "top")


# growth response to light, mechanistic vs hybrid, for four well-sampled species
growth_light <- merge(growth[var == "light" & species %in% c(5, 2, 1, 4)],
                species_dt, by = "species")

ggplot(growth_light, aes(x, ale, colour = model)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~species_name, nrow = 1, scales = "free") +
  scale_colour_manual(values = c("Process (mechanistic)" = "#1b9e77",
                                 "Hybrid (growth = NN)"  = "#d95f02")) +
  labs(x = "light", y = "accumulated local effect on growth",
       colour = NULL) +
  theme_minimal() + theme(legend.position = "top")

```

For the pines the network recovered the mechanistic decline of growth
with size almost exactly; for *Abies grandis* and *Pseudotsuga* it
settled on a different, data-driven shape — the freedom that makes
hybrids useful when the mechanistic form is uncertain. Each curve spans
only the range its *own* model actually simulates (ALE never
extrapolates), so where the process and hybrid cover different ranges —
as for *Abies grandis* light — the two models simply place that species
in different conditions; the mismatch is information, not an artefact.
The same `ale_*$growth` tables also carry the responses to light and
climate, and `ale_*$mortality` / `$regeneration` the other two processes
— plot any of them the same way. The companion article *Recovering a
forest process with Hybrid-FINN* runs this comparison against a known
simulated truth.

```{r}
#| label: session
sessionInfo()
```
