---
title: "Plausible succession from a handful of species"
format:
  html:
    toc: true
    toc-depth: 3
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{Plausible succession from a handful of species}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

FINN encodes regeneration, growth, mortality and competition as ecological
processes. Here we give it nothing but four species with contrasting
life-history strategies and let it simulate succession on bare ground — to show
that ecologically sensible dynamics emerge from the demographic trade-offs
alone. No data, no fitting.

```{r}
#| label: libraries
library(FINN)
library(data.table)
library(ggplot2)
FINN.seed(1234)

Ntimesteps <- 300
Nsites     <- 1
Nsp        <- 4          # pioneer -> early-mid -> mid-late -> climax
patch_size <- 0.1
```

## The species and their trade-offs

We place four species along a classic life-history axis, from a fast,
light-demanding **pioneer** to a slow, shade-tolerant **climax** species. Every
parameter below encodes one end of a trade-off.

The single most important lever is the **shade parameter** (`parGrowth[,1]` and
`parReg`). In FINN this is the fraction of light a species needs: growth and
regeneration are gated by a sigmoid centred on `light = shade`, so a species
only performs where `light > shade`. A **high** value therefore means
*light-demanding* (a pioneer that needs open ground); a **low** value means
*shade-tolerant* (a climax species that establishes under a closed canopy).

```{r}
#| label: species-parameters
# shade = light fraction a species needs to grow / regenerate
#   high -> light-demanding (pioneer); low -> shade-tolerant (climax)
shadeSP <- c(0.60, 0.42, 0.25, 0.08)   # sp1 pioneer ... sp4 climax

# --- regeneration ----------------------------------------------------------
# light threshold tracks shade tolerance; overall seed rain set by the
# regeneration intercept (pioneer prolific, climax sparse but persistent).
parReg    <- shadeSP
parRegEnv <- list(matrix(c(
  c(3.2, 2.6, 2.0, 1.2),   # intercept: overall regeneration effect size
  rep(0, Nsp)              # env slope = 0 (we hold the environment constant)
), Nsp, 2))

# --- growth ----------------------------------------------------------------
# col 1 = shade threshold (as above); col 2 = size-dependent growth decay.
parGrowth <- matrix(c(
  shadeSP,
  c(0.040, 0.038, 0.034, 0.030)
), Nsp, 2)
# growth-rate intercept: the pioneer grows ~4x faster than the climax, so it
# seizes the early canopy; the climax builds slowly.
parGrowthEnv <- list(matrix(c(
  c(1.00, 0.65, 0.30, -0.20),
  rep(0, Nsp)
), Nsp, 2))

# --- mortality -------------------------------------------------------------
# Mortality is mort = plogis(intercept + col1*light + col3*growth), so it
# changes through the run as light and growth change:
#   col1 (light): negative -> mortality RISES as a cohort is overtopped (low
#                 light). Steep for the pioneer (shade-intolerant), ~flat for
#                 the climax (shade-tolerant).
#   col3 (growth): negative -> suppressed, slow-growing trees die more (vigour).
#   col2 (dbh): left 0.
parMort <- matrix(c(
  c(-1.0, -0.8, -0.5, -0.3),   # col1 light: pioneer strongly shade-intolerant
  rep(0, Nsp),                 # col2 dbh: off
  c(-0.3, -0.3, -0.3, -0.3)    # col3 growth: vigour lowers mortality
), Nsp, 3)
parMortEnv <- list(matrix(c(
  c(-0.69, -1.40, -2.30, -3.30),  # intercept: per-species baseline level
  rep(0, Nsp)
), Nsp, 2))

# --- competition -----------------------------------------------------------
# col 1 = height allometry (the climax is tallest and overtops late);
# col 2 = competition strength.
parComp <- matrix(c(
  c(0.50, 0.55, 0.62, 0.70),
  c(0.30, 0.25, 0.20, 0.15)
), Nsp, 2)
```

Collected into one table, the trade-offs read across the life-history axis:

```{r}
#| label: pars-table
pars_dt <- data.table(
  species       = c("1 pioneer", "2 early-mid", "3 mid-late", "4 climax"),
  shade         = shadeSP,
  growth_rate   = round(exp(parGrowthEnv[[1]][, 1]), 2),   # relative growth speed
  mort_shaded   = round(plogis(parMortEnv[[1]][, 1]), 3),  # mortality when overtopped (light~0)
  shade_intol   = -parMort[, 1],                           # how much mortality rises in shade
  seed_rain     = round(exp(parRegEnv[[1]][, 1]), 1),      # relative fecundity
  height_allom  = parComp[, 1]
)
pars_dt
```

Reading the rows: the pioneer needs the most light (`shade` 0.60), grows fastest
(`growth_rate` ≈ 2.7), is the most shade-intolerant (`shade_intol` highest, so its
mortality climbs steeply once overtopped — `mort_shaded` ≈ 0.33) and seeds most
prolifically; the climax is the mirror image — shade-tolerant, slow, almost
immortal even in deep shade (`mort_shaded` ≈ 0.035), sparse but steady — and the
tallest, so it ultimately overtops.

## A constant environment

To isolate the trade-offs we hold the environment constant (`env1 = 0`) and run
primary succession from bare ground with no disturbance — so any dynamics come
from the species themselves, not from the environment.

```{r}
#| label: environment
env_dt <- data.table(expand.grid(siteID = 1:Nsites, year = 1:Ntimesteps))
env_dt$env1 <- 0          # constant -> succession is driven by the traits
```

## Simulate succession

Building and simulating the model needs a torch backend, so the call below
is shown but not run when the page is knit; its result is precomputed
(`dev/precompute_vignettes.R`) and loaded from a small cache.

```{r}
#| label: simulate
#| eval: false
m <- finn(
  N_species            = Nsp,
  competition_process  = createProcess(~0,      FINN::competition),
  growth_process       = createProcess(~1+env1, FINN::growth,       initSpecies = parGrowth, initEnv = parGrowthEnv),
  regeneration_process = createProcess(~1+env1, FINN::regeneration, initSpecies = parReg,    initEnv = parRegEnv),
  mortality_process    = createProcess(~1+env1, FINN::mortality,    initSpecies = parMort,   initEnv = parMortEnv)
)

sim <- predict(m, init_cohort = NULL, env = env_dt, disturbance = NULL,
               patches = 100, device = "cpu")

# average over patches; convert per-patch ba/trees to per-hectare
p_dat <- sim$long$site[, .(value = mean(value)), by = .(year, species, variable)]
p_dat[variable %in% c("ba", "trees"), value := value / patch_size]

sp_labs <- c("1 pioneer", "2 early-mid", "3 mid-late", "4 climax")
```

```{r}
#| label: simulate-cache
#| include: false
.vig    <- readRDS(system.file("extdata", "vig_succession.rds", package = "FINN"))
p_dat   <- .vig$p_dat
sp_labs <- .vig$sp_labs
```

## The successional pattern

```{r}
#| label: plot-succession
#| fig-cap: "Basal area by species over time. The pioneer peaks within the first decade and collapses; mid-seral species crest in turn; the shade-tolerant climax rises slowly and dominates the mature stand."
#| fig-width: 7
#| fig-height: 4.5
ba <- p_dat[variable == "ba"]
ggplot(ba, aes(year, value, colour = factor(species))) +
  geom_line(linewidth = 0.9) +
  scale_colour_viridis_d(name = "Species", labels = sp_labs) +
  labs(x = "Year", y = expression(Basal~area~(m^2/ha))) +
  theme_minimal()
```

The same run, across every state variable FINN tracks — diameter, basal area,
tree numbers, growth, mortality and regeneration:

```{r}
#| label: plot-all-variables
#| fig-cap: "Succession across all FINN state variables for the four species."
#| fig-width: 8
#| fig-height: 9
panel <- copy(p_dat)
# regeneration is shown as the mean per hectare (r_mean_ha); the per-patch
# count (reg) is dropped.
panel[, variable2 := factor(
  variable,
  levels = c("dbh", "ba", "trees", "AL", "growth", "mort", "r_mean_ha"),
  labels = c("avg. DBH [cm]", "Basal Area [m2/ha]", "Trees [N/ha]",
             "Available Light [%]", "Growth [cm/yr]", "Mortality [%]",
             "Regeneration [N/ha]"))]
ggplot(panel[!is.na(variable2)], aes(year, value, colour = factor(species))) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~variable2, scales = "free_y", ncol = 2, strip.position = "left") +
  coord_cartesian(ylim = c(0, NA)) +
  scale_colour_viridis_d(name = "Species", labels = sp_labs) +
  labs(x = "Year", y = NULL) +
  theme_minimal() +
  theme(strip.placement = "outside",
        strip.text.y.left = element_text(angle = 90),
        axis.title.y = element_blank())
```

## What just happened

The figure traces textbook succession, and every feature maps back to a
parameter:

- **The pioneer (species 1) erupts and collapses.** On open ground light is
  abundant, so its high light requirement is met; its fast growth and prolific
  seeding let it seize the canopy within the first decade. But as the canopy
  closes, light drops below its high `shade` threshold — it can no longer
  regenerate — and because it is shade-intolerant its mortality climbs steeply
  once overtopped (visible in the mortality panel: from ~0.08 yr⁻¹ in full light
  to ~0.3 yr⁻¹ in shade), clearing the standing trees. By mid-succession it is
  essentially gone.
- **The early- and mid-seral species crest in turn.** Species 2 and 3 have
  progressively lower light requirements and lower mortality, so each peaks a
  little later and lingers a little longer. The mid-late species holds a stable
  subdominant presence through the middle of the run.
- **The climax (species 4) inherits the stand.** It grows slowly and seeds
  sparsely, so early on it is barely visible. But it regenerates under shade
  (lowest `shade`), is shade-tolerant so it barely dies even when overtopped
  (mortality stays ~0.02–0.03 yr⁻¹ throughout), and is the tallest (highest
  `height_allom`) — so once the early-seral canopy senesces it overtops
  everything and dominates the mature forest.

None of this was prescribed: we set demographic trade-offs for four species and
FINN produced the successional turnover.

```{r}
#| label: session
sessionInfo()
```
