---
title: "Recovering a forest process with Hybrid-FINN"
author: "FINN team"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    fig-width: 7
    fig-height: 4.5
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{Recovering a forest process with Hybrid-FINN}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# Heavy training chunks are guarded with eval flags so the article can be
# rendered for the website without a GPU. Set to TRUE to run for real.
RUN_FITS <- FALSE
```

::: {.callout-note appearance="simple"}
**Abstract.** This article shows a complete Hybrid-FINN workflow: we
simulate forest succession from a known process model, then ask whether a
hybrid model—where one ecological process is replaced by a neural
network—can recover the *true functional form* of that process from data
alone. This mirrors the niche-recovery result in Pichler & Käber (2026)
and is the fastest way to see what FINN does end to end.
:::

## The idea

Dynamic forest (gap) models encode ecological processes—regeneration,
growth, mortality—as fixed mathematical forms whose parameters are
calibrated to data. FINN keeps that structure but lets you swap any single
process for a neural network and fit everything jointly by gradient
descent. If the network can re-learn a relationship we deliberately built
into the data-generating model, we have evidence that FINN can also
*discover* relationships we don't know in advance—such as a species'
environmental niche.

We proceed in four steps, just like a species distribution workflow:

1. **Simulate** truth from a Process-FINN with a known environmental niche.
2. **Fit** two competitors: a misspecified process model and a Hybrid-FINN.
3. **Diagnose** convergence and predictive performance.
4. **Infer** the recovered niche with accumulated local effects (ALE).

```{r}
#| label: libraries
library(FINN)
library(data.table)
library(ggplot2)
FINN.seed(42)
```

## 1. Simulate truth

We define five species and a single environmental gradient, give the
growth process a unimodal (quadratic) response to the environment, and
simulate succession forward. The quadratic term is the "true niche" we
will later try to recover.

```{r}
#| label: simulate
#| eval: !expr RUN_FITS
Ntimesteps <- 500
Nsites     <- 50
Nsp        <- 5

# --- species parameters (see Introduction-to-FINN for the full layout) ---
# TODO: confirm exact argument shapes against your installed FINN version.
shadeSP <- c(0.1, 0.2, 0.5, 0.5, 0.7)

parGrowth    <- matrix(c(shadeSP, c(0.04, 0.05, 0.05, 0.06, 0.10)), Nsp, 2)
# unimodal environmental niche on growth: intercept + linear + quadratic
parGrowthEnv <- list(matrix(c(
  runif(Nsp,  0.2, 0.6),   # overall effect size
  runif(Nsp, -1.0, 1.0),   # linear env term
  runif(Nsp, -2.0, -0.5)   # quadratic env term  <- the "niche"
), Nsp, 3))

# (regeneration / mortality parameters omitted here for brevity)

truth <- simulateForest(
  Ntimesteps = Ntimesteps,
  Nsites     = Nsites,
  # ... process + environment inputs ...
)
```

```{r}
#| label: plot-succession
#| eval: !expr RUN_FITS
#| fig-cap: "Simulated basal area by species over time at one site."
df <- pred2DF(truth)
ggplot(df[df$siteID == 1, ], aes(year, ba, colour = factor(species))) +
  geom_line(linewidth = 0.8) +
  scale_colour_viridis_d(name = "Species") +
  labs(x = "Year", y = expression(Basal~area~(m^2))) +
  theme_minimal()
```

## 2. Fit two competitors

We turn the simulated trajectories into an observation table and fit:

- **M2 — misspecified Process-FINN:** growth–environment assumed *linear*.
- **M3 — Hybrid-FINN:** growth's environmental response replaced by a DNN.

```{r}
#| label: build-obs
#| eval: !expr RUN_FITS
obs <- makeObsData(truth)        # observation table FINN fits against
```

```{r}
#| label: fit-models
#| eval: !expr RUN_FITS
# M2: process model, linear environment (misspecified on purpose)
m2 <- finn( /* growth = createProcess(... linear env ...) */ )
m2 <- fit(m2, data = obs, epochs = 150L, lr = 0.01)

# M3: hybrid — growth's env response is a neural network
growth_nn <- createHybrid( /* hidden = c(20, 20), env -> growth effect */ )
m3 <- finn( /* growth = growth_nn */ )
m3 <- fit(m3, data = obs, epochs = 150L, lr = 0.01)
```

::: {.callout-tip}
Training neural networks inside a dynamic model is finicky—watch for a
still-decreasing loss (train longer / raise the learning rate) and jumpy
late-epoch loss (add weak regularization). The same convergence caveats
that apply to plain DNNs apply here.
:::

## 3. Diagnose

Compare predicted vs. observed at equilibrium with a Spearman correlation,
the metric used in the paper.

```{r}
#| label: performance
#| eval: !expr RUN_FITS
pred_m2 <- pred2DF(m2)
pred_m3 <- pred2DF(m3)
# cor(pred, obs, method = "spearman") for ba / trees / dbh ...
```

## 4. Infer the recovered niche

The payoff: does the hybrid's learned growth–environment curve match the
true quadratic niche? Accumulated local effects (ALE) read the answer out
of the fitted model—exactly as `ALE()` does for a `cito` DNN.

```{r}
#| label: ale
#| eval: !expr RUN_FITS
#| fig-cap: "Recovered growth-environment response (M3, hybrid) vs. the misspecified linear model (M2) and the true niche."
ALE(m3, variable = "env")   # add M2 + truth overlays for the comparison figure
```

If M3's curve traces the true unimodal niche while M2's straight line
cannot, FINN has recovered the functional form from data alone.

## Where to go next

- Swap the simulated data for real inventory data (see the *Fit FINN to
  FIA* article) to infer **empirical** species niches.
- Replace a different process (regeneration, mortality) and compare which
  is most identifiable.

## Session info

```{r}
#| label: session
sessionInfo()
```
